Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-22 aistats.tex
Add Gaussian figures

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
We thus consider the fundamental problem: %TODO thus
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

@@

In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
% TODO We also give complementary illustrations on the Gaussian mirror map and MLE.

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
% TODO These functions and their domains are represented in Figure \TODO{}.
We provide definitions of self-concordance in \cref{app:self-concordant}.

@@

\section{BIAS-VARIANCE}
\label{app:bias-variance}
In this section, we start from the notions of bias and variance introduced in \cref{eq:bias-variance}.
First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields to a convergence rate on the variance term.

@@

\subsection{Expectation of SMD's Variance Assumption}
The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
\alignn{
	\cS_\conj (\mu , \bar\mu)
	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
}
Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
This means that
\alignn{
	\expect{\cS_\conj (\mu , \E[\mu])} 
	= \expect{\lin{\nat, \mu - \E[\mu]}}
	= \underbrace{\expect{\lin{\nat, \mu}} -\lin{\expect{\nat}, \expect{\mu} } }_{\Cov(\nat, \mu)}
	=\expect{\lin{\nat - \E[\nat],\mu}} 
	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
}
The last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
Now, \cref{eq:bias-variance} defines the variance as 
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
	\label{eq:bregman-covariance}
}
where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
\alignn{
	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
}
where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
Taking the expectation over $\hat\mu_{n-1}$ instead gives
\alignn{
	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
}
The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
\alignn{
	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} },
}
while
\alignn{
	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
}
In the end, we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}:
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
}
If assumption~\eqref{eq:hanzely} holds, then we have
\alignn{
\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
}
where we assumed $\lr_n  \in O\paren{\inv{n}}$.
In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.

>>>>

\paragraph{Statistical decision theory}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
We thus consider the fundamental problem: %TODO thus repetition
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

@@

In this section, we prove that for a Gaussian, the entropy and log-partition functions are self-concordant.
We also provide complementary illustrations of 
these functions in \cref{fig:logpart-entropy}, 
their gradients (e.g. the mirror maps) in \cref{fig:mirrormaps},
and paths taken by MLE and MAP in \cref{fig:gaussian-paths} .

The Gaussian log-partition function and entropy are, up to constants,
\alignn{
	\logpart(\nat) &= \frac{\nat_1^2}{-4\nat_2} - \half \log(-\nat_2) \label{eq:app-logpart} \\
	\conj(\m) &= - \half \log (\mu_2 - \mu_1^2) \; , \label{eq:app-entropy}
}
where $\nat_1\in\real, \nat_2 <0$ and $\mu_1\in\real, \mu_2>\mu_1^2$.
We provide definitions of self-concordance in \cref{app:self-concordant}.

\begin{figure}[ht]
	\centering
	\includegraphics[width=.45\textwidth]{figs/2d/logpart.pdf}
	\includegraphics[width=.45\textwidth]{figs/2d/entropy.pdf}
	\caption{
	(left) Contours of the log-partition function~\eqref{eq:app-logpart}.
	(right) Contours of the entropy~\eqref{eq:app-entropy}.
	}
	\label{fig:logpart-entropy}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{figs/2d/mirrormap.pdf}
	\caption{
	(top) Grid deformation produce by $\nabla \logpart(\nat)$.
	(bottom) Grid deformation produced by $\nabla\conj(\m)$.
	}
	\label{fig:mirrormaps}
\end{figure}

\begin{figure}[ht]
	\centering
	\includegraphics[width=.9\textwidth]{figs/2d/gaussian_paths.pdf}
	\caption{
		Paths taken by MLE (blue) and MAP (orange) on top of contours for $\KL(\nat^* || \nat)$. 
		We set $\mu^*=(0, 1), \mu_0 = (1, 2)$, and $n_0=4$, and $n$ varies from $1$ to $20$.
		In green, red and purple, we represent the paths respectively taken by the MAP dual expectation  $(\bar\nat_n, \bar \mu_n)$, MAP primal expectation $(\tilde\nat_n,\tilde\mu_n)$, and MLE primal expectation. Recall that the MLE dual expectation is $\mu^*$ itself.
	}
	\label{fig:gaussian-paths}
\end{figure}

@@

\section{BIAS-VARIANCE}
\label{app:bias-variance}
In this section, we start from the notions of bias and variance introduced in \cref{eq:bias-variance}.
First, we prove that the bias of the MLE of a Gaussian variance decreases in $O(\inv{n^2})$. 
Then we prove that assuming~\eqref{eq:hanzely} holds, whether uniformly or in expectation, yields a convergence rate on the variance term.

@@

\subsection{Expectation of SMD's Variance Assumption}
The first step is to notice the symmetrized Bregman can be expressed as an inner product between primal and dual parameters
\alignn{
	\cS_\conj (\mu , \bar\mu)
	= \lin{ \nabla \conj(\mu) - \nabla \conj(\bar\mu) , \mu - \bar\mu}
	= \lin{ \nat -\bar\nat , \mu - \bar\mu} \; .
}
Now notice that~\eqref{eq:hanzely} features $\mu := \hat \mu_{n+1} = \hat\m_{n} - \lr g(\hat \nat_{n})$ stochastic and $\bar\mu := \bar\mu_{n+1} =  \hat\m_{n}   -\lr \nabla f(\hat \nat_{n})$ deterministic such that $\E_g[\mu] = \bar\mu$.
For such a pair of variables, the expectation of the symmetrized Bregman corresponds to a covariance between primal and dual parameters
\alignn{
	\expect{\cS_\conj (\mu , \E[\mu])} 
	= \expect{\lin{\nat, \mu - \E[\mu]}}
	= \underbrace{\expect{\lin{\nat, \mu}} -\lin{\expect{\nat}, \expect{\mu} } }_{\Cov(\nat, \mu)}
	=\expect{\lin{\nat - \E[\nat],\mu}} 
	= \expect{\cS_\logpart(\nat, \E[\nat])} \; .
}
The last equality holds by symmetry between the roles of $\logpart$ and $\conj$.
Note that the middle covariance formulation is actually the one  used by \citet{hanzely2018fastest}.
Now, \cref{eq:bias-variance} defines the variance as 
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	=\expect[1:n]{\bregman(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	\leq \expect[1:n]{\cS_A(\hat\nat_n , \E_{1:n}[\hat\nat_n]) }
	=  \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }\;,
	\label{eq:bregman-covariance}
}
where expectations $\E_{1:n}$ are on all samples $X_1, \dots, X_n$, whereas~\eqref{eq:hanzely} is written with 
\alignn{
	\expect[n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) } \leq \lr^2 C \; ,
}
where the expectation $\E_{n}$ is taken over only the last sample $X_n$, and the bound should hold uniformly over all $\hat\mu_{n-1}$.
Taking the expectation over $\hat\mu_{n-1}$ instead gives
\alignn{
	\expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \leq \lr^2 C \;.
}
The only difference with the RHS of \cref{eq:bregman-covariance} is in the inner expectation.
To overcome this difference, we need to plug in the form of $\hat \mu_n = \frac{n_0\mu_0 + \sum_i T_i}{n_0 +n}$. Notice that 
\alignn{
	&\MAPm - \E_n[\MAPm] = \frac{T_n - \m^*}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }
	= \inv{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} },
}
while
\alignn{
	&\MAPm - \E_{1:n}[\MAPm] = \frac{\smallsum_i (T_i - \m^*)}{n_0+n} \\
	\implies & \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{1:n}[\hat\mu_n]) }
	=\frac{n}{n_0+n} \expect[1:n]{ \lin{\MAPt , T_1 - \m^*} } \; .
}
In the end, we get that the variance is dominated by $n$ times the expectation of \cref{eq:hanzely}:
\alignn{
	\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
	\leq n \expect[1:n]{\cS_\conj(\hat\mu_n , \E_{n}[\hat\mu_n]) }  \; .	
}
If assumption~\eqref{eq:hanzely} holds, then we have
\alignn{
\expect[1:n]{\bregmanconj(\tilde \m_n , \hat\m_n) }
\leq n \lr_n^2 C \in O\paren{\inv{n}} \; ,
}
where we assumed $\lr_n  \in O\paren{\inv{n}}$.
In conclusion, assuming~\eqref{eq:hanzely} holds uniformly or in expectation immediately implies a $O\paren{\inv{n}}$ convergence rate on the variance.
